---
title: "The effect of COVID-19 pandemic on Croatian tourist sector"
author: "Bogdan, S., Sikic, L. and Suzana,XXX"
date: '`r format(Sys.time(), "%d %B, %Y")`'
header-includes:
  - \usepackage{caption}
  - \captionsetup[figure]{labelformat=empty}
output:
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::pdf_document2:
    includes:
      before_body:
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: true
    toc: no
  bookdown::word_document2: null 
bibliography: "D:/LUKA/Academic/Covid-vs-Tourism/Literatura/reference.bib"
fontsize: 12pt
link-citations: yes
---

```{r library, warning=FALSE, include=FALSE, message=FALSE}

library(estudy2)
library(plotly)
library(TSstudio)
library(jsonlite)
library(plyr)
library(zoo)
library(httr)
library(plotly)
library(jsonlite)
library(dplyr)
library(tidyr)
library(Quandl)
library(stringr)
library(eventstudies)
library(stats)
library(kableExtra)
        
```

# INTRODUCTION

The coronavirus pandemic flooded the globe in just a few months after the first case was registered, leaving huge consequences in the form of threatening the health and human lives, great economic losses, and psychological fear that has crept deep into society. The presence of the virus around the world has prevented potential tourists from feeling safe in the destination which has left deep consequences today and in the future for one of the most affected sectors, and that is tourism. According to the UNWTO (2021) international tourist arrivals (overnight visitors) dropped by 74% in 2020 compared to the 2019, due to massive drop in tourist demand and travel restrictions. This pandemic has caused loss of USD 1.3 trillion in export revenues which is eleven times greater loss compared to the global financial crisis from 2008. UNWTO forecasted that it could pass between 2,5 or even 4 years for international tourism to return on the old track from 2019. The effectiveness of vaccines and the level of vaccination of the population will certainly play an important role, which should contribute to a smaller number of new cases, as well as to the mitigation of travel restriction measures and the restoration of consumer confidence. The consequences are even worse since Croatia is a highly tourism dependent country.  

According to the UNWTO, Croatia is among the 10 most vulnerable countries according to the criterion of the direct impact of tourism on the share of GDP. More precisely, in the first place is Macao (China) 48%, followed by: Fiji 13%, Jordan 12%, Spain 12% and Croatia with 11%. Many countries have introduced travel restrictions, and knowing that the share of foreign tourists in Croatia is 89% (UNWTO, 2021), negative results are expected from companies that are directly or indirectly involved in tourism. In Croatia tourism is taking part in a large share of total exports (35%) as well (UNWTO, 2021). Although Croatia achieved a record in 2019 in terms of the total number of arrivals and overnight stays of domestic and foreign tourists in 2020 due to the global corona pandemic, there was a steep decline compared to 2019, which amounts to -55,29% of overnight stays and -64,22% of total arrivals. Poor results in tourism spilled over into the capital market and caused turmoil, so the purpose of this research was to investigate how tourism stock prices respond to the event of the COVID-19 pandemic by using event study technique. This study contributes to the literature by analyzing the impact of global COVID-19 pandemic on tourism stock volatility in Croatia. Tu još nešto fali. The remainder of this study is structured as follows……… 



# RESEARCH BACKGROUND AND LITERATURE REVIEW

Tourist indicators of Croatian tourism before and during the Covid-19 pandemic 

 

The Republic of Croatia systematically follows modern tourist trends and is well positioned on the European tourist market. It is recognized as a stable, safe tourist destination, beautiful and rich natural and cultural-historical heritage. The Croatian tourism sector has been successful since Croatia's independence and accession to the European Union, and has consistently recorded enviable results, until the outbreak of the coronavirus pandemic in 2020. According to the data published on the official website of the World Health Organization–WHO, by April 20, 2021, a total of 141,754,944 confirmed cases of COVID-19 were recorded, of which 3,025,835 deaths. According to the data reported to the WHO, from 3 January 2020 to 20 April 2021, 310,306 confirmed COVID-19 cases with 6,643 deaths were recorded in Croatia on the total population of 4,058,165 

The SARS-CoV-2 virus pandemic has shaken the whole world and caused historically unique problems in all aspects of social and economic life. It has greatly affected the global economic activities. Measures taken to prevent virus spreading (social distancing, traffic restrictions, restrictions on commercial activity, borders closures, etc.) have affected in particular the service sector, tourism and hospitality. The analysis of selected indicators of tourism development in Croatia before and at the time of Covid-19 pandemic in Croatia, shows how much the pandemic affected the results in tourism. 

 


According to the historical review of data on tourist arrivals and overnight stays in Croatia (table 1), a constant growth in tourist arrivals and tourist overnight stays can be observed until 2020. If we compare the nights and arrivals in 2019 with those from 2010, it could be stated that the results have almost doubled. Despite the favorable results, a sharp decline followed in 2020 with decrease in tourist arrivals by 64.2% and decrease in overnight stays by 55.3% in commercial accommodation in 2020 compared to 2019 year. The decline in arrivals and overnight stays followed among domestic and foreign tourists, with the same being more pronounced among foreign tourists, which can also be seen from Charts 1 and 2. 

Tu fali uvodni dio. 

@Pany used a sample of 46 stocks listed in the Stock Exchange of Thailand in order to analyze stock price reactions during pandemic. After applying event study methodology in order to empirically measure abnormal returns and volatility, research results suggest that the majority of stocks in the Stock Exchange of Thailand have been negatively affected by Covid-19 pandemic. On the other side authors have also found that some stocks had positive returns, these stocks are included in commerce sector- companies which are included in distribution of pharmaceutical products and services. Irfan, Kassim and Dhimmar (2021) analyzed the impact of COVID-19 on the performance of Indian stock exchange and Indonesian Stock Exchange. Author included three different event windows because they wanted to check results in different time periods. The analysis of both countries showed opposite results since Indian stock market showed downward sloping after WHO declaration, and Indonesian upward. @Abul researched new COVID-19 official cases announcements and death ratio on the financial markets volatility index (VIX). Authors has concluded that new cases reported outside China has positive impact on VIX, death ratio has significant positive impact on VIX, and the spread of the COVID-19 increase financial volatility. Chia, Khim-Sen Liew and Rowland (2020) research the relation between Malaysian stock market and variables related to COVID-19 their results suggest that daily new cases had negative but insignificant impact on the indices returns. On the other hand authors also found that movement control order had significant and positive impact on all indices’ returns which is quite surprising.  

 

# DATA COLLECTION AND METHODOLOGY

```{r importData, echo=FALSE,  message = FALSE, warning = FALSE}

ISIN <- readxl::read_xlsx("./Data/ISIN_list.xlsx") %>% select(ISIN) 

links <- c()
rawDta <- c()
for( i in ISIN){ 
  links <- paste0('https://zse.hr/json/securityHistory/', i,
                  '/2019-01-01/2021-04-13/hr?trading_model_id=ALL')
}
for (i in links){
rawDta[[i]] <- fromJSON(content(GET(i), as = "text", encoding = "UTF-8")) 
}

stockDF <- lapply(rawDta, '[[', "rows") %>% bind_rows()


source("./Secret/Passkey.R")
Crobex <- Quandl("ZAGREBSE/CROBEX", type = "zoo",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date())

gold <- Quandl("LBMA/GOLD", type = "zoo", collapse = "daily", start_date = "2019-01-01",end_date = Sys.Date())
gold <- gold[,5]


options(scipen = 999,pillar.sigfigs=3)
CrobexKonstr <- Quandl("ZAGREBSE/CROBEXKONSTRUKT", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexKonstr = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexInd <- Quandl("ZAGREBSE/CROBEXINDUSTRIJA", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexInd = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexNutr <- Quandl("ZAGREBSE/CROBEXNUTRIS", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexNutr = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexTur <- Quandl("ZAGREBSE/CROBEXTURIST", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexTur = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexTr <- Quandl("ZAGREBSE/CROBEXTR", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexTr = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

market <- CrobexKonstr %>%
  left_join(.,CrobexInd, by = "Date") %>%
  left_join(.,CrobexNutr, by = "Date") %>%
  left_join(.,CrobexTur, by = "Date") %>%
  left_join(.,CrobexTr, by = "Date") %>%
  gather( Index,Return, -Date)

#range(market$Date)
```

```{r cleanData, echo=FALSE,  message = FALSE, warning = FALSE}

TOURISMdta <- stockDF %>% select(Date = date, Close = last_price, url) %>%
  mutate(Date = gsub("[.]$","", Date)) %>%
  mutate(Ticker = str_sub(.$url,-37,-34)) %>%
  mutate(Close = as.numeric(gsub("(.*),.*", "\\1", Close))) %>%
  select(-url) %>%
  group_by(Ticker) %>%
  dplyr::mutate(n = n()) %>%
  ungroup() %>% 
  dplyr::filter(n > 100) %>%
  dplyr::select(-n) %>%
  group_by(Ticker) %>%
  distinct(Date, .keep_all=TRUE) %>%
  ungroup() %>%
  tidyr::spread(Ticker, Close, fill=0) %>%
  mutate( Date = as.Date(Date,"%d.%m.%Y")) %>%
  arrange(desc(Date))


TOURISMdta <- zoo(TOURISMdta[,-1], order.by = TOURISMdta$Date)


 market %>%
  mutate(dr = cut(Date, breaks = c(range(Date), 
            as.Date(c("2020-02-12", "2020-04-08","2021-04-15"))), include.lowest=TRUE)) %>% #View()
  mutate(Pandemic =  case_when(
    .$dr == "2019-01-03" ~ "Pre",
    .$dr == "2020-02-12" ~ "Ongoing",
    .$dr == "2020-04-08" ~ "Post")) %>%  #
  filter(Date < "2021-04-15") %>% 
  select(-dr) %>% arrange (Date) %>%  #
  group_by(Pandemic, Index) %>%
  summarise(Return_sum = sum(Return),
            Return_avg = mean(Return),  
            Return_sd = sd(Return),
            Return_min = min(Return),
            Return_max = max(Return)) %>% 
  mutate_if(is.numeric, round, 3) %>% kable()

```

```{r stockReturns, echo=FALSE, message=FALSE, warning=FALSE}
rates <- get_rates_from_prices(TOURISMdta,
                               quote = "Close",
                               multi_day = TRUE,
                               compounding = "continuous")

ratesDF <- data.frame(rates)
ratesDF <- data.frame(Date = rownames(ratesDF), rates) %>% mutate(Date = as.Date(Date,"%Y-%m-%d"))
rownames(ratesDF) <- NULL
ratesDF[sapply(ratesDF, is.infinite)] <- NA
ratesDF[sapply(ratesDF, is.nan)] <- NA
ratesDF[sapply(ratesDF, is.na)] <- 0
rates <-  zoo(ratesDF[,-1], order.by = ratesDF$Date)
```

```{r crobexReturn, echo=FALSE, message=FALSE, warning=FALSE}
rates_indx <- get_rates_from_prices(Crobex, 
                                    quote = "Close",
                                    multi_day = TRUE,
                                    compounding = "continuous")

colnames(rates_indx) <- "Crobex"
```

```{r goldReturn, echo=FALSE, message=FALSE, warning=FALSE}
add_indx <- get_rates_from_prices(gold, 
                                  quote = "Close",
                                  multi_day = TRUE,
                                  compounding = "continuous")

colnames(add_indx) <- "gold"
```

## Methodology


Data and Methodology 

 

This research paper will study the impact of variables related to the COVID 19 pandemic on the shares of the tourism sector listed on the Croatian capital market. The study used daily closing stock prices in order to calculate daily percentage returns. However, since the Croatian capital market has major liquidity problems out of a total of XX shares listed on the Croatian capital market, in this study was used XX shares according to the criterion of the number of trading days, which had to be greater than 100. 

 

Daily percentage changes of closing stock prices were calculated according to formula: 

## Insigthts from the data and topic identification

 


# RESULTS AND DISCUSSION


## Nonparametric tests

```{r, echo=FALSE, message=FALSE, warning=FALSE}
securities_returns <- apply_market_model(
  rates = rates,
  regressor = rates_indx,
  same_regressor_for_all = TRUE,
  market_model = "sim",
  estimation_method = "ols",
  estimation_start = as.Date("2019-01-01"),
  estimation_end = as.Date("2020-01-23")
)

parametric_tests(list_of_returns = securities_returns,
                 event_start = as.Date("2020-02-24"),
                 event_end = as.Date("2020-04-14"))
```



## Parametric tests

```{r, echo=FALSE, message=FALSE, warning=FALSE}
nonparametric_tests(list_of_returns = securities_returns,
                    event_start = as.Date("2020-02-24"),
                    event_end = as.Date("2020-04-14"))
```


## Visual presentation of results


```{r vizNOexter, echo=FALSE, message=FALSE, warning=FALSE}
SplitD <- data.frame(name=colnames(rates),
                     when = rep(as.Date("2020-02-24"),12))

Returns <- na.omit(rates)

es <- eventstudy(firm.returns = Returns,
                 event.list = SplitD,
                 event.window = 40,
                 type = "None",
                 to.remap = TRUE,
                 remap = "cumsum",
                 inference = TRUE,
                 inference.strategy = "bootstrap")

plot(es)
```

```{r vizCROBEXexter, echo=FALSE, message=FALSE, warning=FALSE}
es2 <- eventstudy(firm.returns = Returns,
                 event.list = SplitD,
                 event.window = 40,
                 type = "marketModel",
                 to.remap = TRUE,
                 remap = "cumsum",
                 inference = TRUE,
                 inference.strategy = "bootstrap",
                 model.args = list(market.returns=rates_indx$Crobex))

plot(es2)
```

```{r vizGOLDexter, echo=FALSE, message=FALSE, warning=FALSE}
es3 <- eventstudy(firm.returns = Returns,
                  event.list = SplitD,
                  event.window = 40,
                  type = "marketModel",
                  to.remap = TRUE,
                  remap = "cumsum",
                  inference = TRUE,
                  inference.strategy = "bootstrap",
                  model.args = list(market.returns=rates_indx$Crobex,
                  others=add_indx$gold,
                  market.returns.purge=TRUE))

plot(es3)
```

# CONCLUSION

# BIBLIOGRAPHY




