---
title: "The effect of COVID-19 pandemic on Croatian tourist sector"
author: "Bogdan,S., Sikic,L. and Baresa,S."
date: "Faculty of Tourism and Hospitality Management, Rijeka"
output:
  xaringan::moon_reader:
    css:
    - default
    - metropolis
    - metropolis-fonts
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: yes
      countIncrementalSlides: no
  html_document:
    df_print: paged
  pdf_document: default
subtitle: Tourism in Southern and Eastern Europe, 6th International Scientific Conference
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```

```{css, echo = F, eval = T}
@media print {
  .has-continuation {
    display: block !important;
  }
}
remark-slide-content {
  font-size: 22px;
  padding: 20px 80px 20px 80px;
}
.remark-code, .remark-inline-code {
  background: #f0f0f0;
}
.remark-code {
  font-size: 16px;
}
.mid. remark-code { /*Change made here*/
  font-size: 60% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 40% !important;
}
```

```{r packgaes, warning=F, echo=F, message=F}

library(estudy2)
library(plotly)
library(TSstudio)
library(jsonlite)
library(plyr)
library(zoo)
library(httr)
library(plotly)
library(jsonlite)
library(dplyr)
library(tidyr)
library(Quandl)
library(stringr)
library(eventstudies)
library(stats)
library(kableExtra)
library(patchwork)
```



# Overview

<br>
<br>
<br>


1. [COVID-19, economy and tourism](#econ)

2. [Data and Event Study metodology](#dtamet)

3. [Effect of COVID-19 on tourist stocks in Croatia](#effect)


<a href="https://your-url" class="github-corner" aria-label="View source on Github"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#fff; color:#151513; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

.footnote[ [*]The analysis is avaliable as [The effect of COVID-19 pandemic on Croatian tourist sector](https://raw.githack.com/lusiki/Covid-vs-Tourism/main/Covid-vs-Tourism-PAPER.html) and all supporting code, data and materials are in the [Github repo](https://github.com/lusiki/Covid-vs-Tourism).

---
class: inverse, center, middle
name: econ

# COVID-19, ECONOMY AND TOURISM

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

(What sector was hit hardest?)

---

# Macroeconomic effects of COVID-19
<br>
<br>

- Huge literature
<br>
- Results of that literature
<br>
- Sectoral effects
<br>



---

# COVID-19 and tourism
<br>
<br>




---
class: inverse, center, middle
name: dtamet

# DATA AND EVENT STUDY METHODOLOGY

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

(The research setup!)

---

# Data from Zagreb Stock Exhange

- Crobex tourist and sectoral indices

- All tourist stocks

---

# Event study methodology

Stock market returns are estimated in the pre-event time period, according to the formula:

$R_{it}=\frac{P_{i,t}-P_{i,t-1}}{P_{i,t-1}}$

where $P_{it}$ is the price of the stock *i* in the period *t*. The abnormal return $AR_{it}$ is defined as a difference between the actual and expected return $E[R_{it}|I_{t}]$ during the event window:

$AR_{it} = R_{it}-E[R_{it}|I_{t}]$.




---
class: inverse, center, middle
name: dtamet

# COVID-19 AND TOURIST STOCKS IN CROATIA

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

(Not that severe effect in tourist sector!)

---

# Market-wide impact


```{r importData, echo=FALSE,  message = FALSE, warning = FALSE}

ISIN <- readxl::read_xlsx("../Data/ISIN_list.xlsx") %>% select(ISIN) 

links <- c()
rawDta <- c()
for( i in ISIN){ 
  links <- paste0('https://zse.hr/json/securityHistory/', i,
                  '/2019-01-01/2021-04-13/hr?trading_model_id=ALL')
}
for (i in links){
rawDta[[i]] <- fromJSON(content(GET(i), as = "text", encoding = "UTF-8")) 
}

stockDF <- lapply(rawDta, '[[', "rows") %>% bind_rows()


source("../Secret/Passkey.R")
Crobex <- Quandl("ZAGREBSE/CROBEX", type = "zoo",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date())

gold <- Quandl("LBMA/GOLD", type = "zoo", collapse = "daily", start_date = "2019-01-01",end_date = Sys.Date())
gold <- gold[,5]


options(scipen = 999,pillar.sigfigs=3)
CrobexKonstr <- Quandl("ZAGREBSE/CROBEXKONSTRUKT", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexKonstr = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexInd <- Quandl("ZAGREBSE/CROBEXINDUSTRIJA", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexInd = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexNutr <- Quandl("ZAGREBSE/CROBEXNUTRIS", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexNutr = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexTur <- Quandl("ZAGREBSE/CROBEXTURIST", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexTur = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

CrobexTr <- Quandl("ZAGREBSE/CROBEXTR", type = "raw",collapse = "daily",start_date = "2019-01-02", end_date = Sys.Date()) %>% 
  arrange(Date) %>%
  mutate(CrobexTr = ((Value - lag(Value))/Value)*100) %>%
  drop_na() %>%
  select(-Value)

market <- CrobexKonstr %>%
  left_join(.,CrobexInd, by = "Date") %>%
  left_join(.,CrobexNutr, by = "Date") %>%
  left_join(.,CrobexTur, by = "Date") %>%
  left_join(.,CrobexTr, by = "Date") %>%
  gather( Index,Return, -Date)

#range(market$Date)
```

```{r cleanData, echo=FALSE,  message = FALSE, warning = FALSE}

TOURISMdta <- stockDF %>% select(Date = date, Close = last_price, url) %>%
  mutate(Date = gsub("[.]$","", Date)) %>%
  mutate(Ticker = str_sub(.$url,-37,-34)) %>%
  mutate(Close = as.numeric(gsub("(.*),.*", "\\1", Close))) %>%
  select(-url) %>%
  group_by(Ticker) %>%
  dplyr::mutate(n = n()) %>%
  ungroup() %>% 
  dplyr::filter(n > 100) %>%
  dplyr::select(-n) %>%
  group_by(Ticker) %>%
  distinct(Date, .keep_all=TRUE) %>%
  ungroup() %>%
  tidyr::spread(Ticker, Close, fill=0) %>%
  mutate( Date = as.Date(Date,"%d.%m.%Y")) %>%
  arrange(desc(Date))


TOURISMdta <- zoo(TOURISMdta[,-1], order.by = TOURISMdta$Date)

```

```{r stockReturns, echo=FALSE, message=FALSE, warning=FALSE}
rates <- get_rates_from_prices(TOURISMdta,
                               quote = "Close",
                               multi_day = TRUE,
                               compounding = "continuous")

ratesDF <- data.frame(rates)
ratesDF <- data.frame(Date = rownames(ratesDF), rates) %>% mutate(Date = as.Date(Date,"%Y-%m-%d"))
rownames(ratesDF) <- NULL
ratesDF[sapply(ratesDF, is.infinite)] <- NA
ratesDF[sapply(ratesDF, is.nan)] <- NA
ratesDF[sapply(ratesDF, is.na)] <- 0
rates <-  zoo(ratesDF[,-1], order.by = ratesDF$Date)
```

```{r crobexReturn, echo=FALSE, message=FALSE, warning=FALSE}
rates_indx <- get_rates_from_prices(Crobex, 
                                    quote = "Close",
                                    multi_day = TRUE,
                                    compounding = "continuous")

colnames(rates_indx) <- "Crobex"
```

```{r goldReturn, echo=FALSE, message=FALSE, warning=FALSE}
add_indx <- get_rates_from_prices(gold, 
                                  quote = "Close",
                                  multi_day = TRUE,
                                  compounding = "continuous")

colnames(add_indx) <- "gold"
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
 market %>%
  mutate(dr = cut(Date, breaks = c(range(Date), 
            as.Date(c("2020-02-12", "2020-04-08","2021-04-15"))), include.lowest=TRUE)) %>% #View()
  mutate(Pandemic =  case_when(
    .$dr == "2019-01-03" ~ "Pre",
    .$dr == "2020-02-12" ~ "Ongoing",
    .$dr == "2020-04-08" ~ "Post")) %>%  #
  filter(Date < "2021-04-15") %>% 
  select(-dr) %>% arrange (Date) %>%  #
  group_by(Pandemic, Index) %>%
  summarise(Total = sum(Return),
            Average = mean(Return),  
            Stdev = sd(Return),
            Min = min(Return),
            Max = max(Return)) %>% 
  mutate_if(is.numeric, round, 3) %>% kable(caption = "Table 1: Overview of the market returns over the analysed period")

```

```{r sectoralEffect, echo=FALSE, message=FALSE, warning=FALSE}
securities_returns <- apply_market_model(
  rates = rates,
  regressor = rates_indx,
  same_regressor_for_all = TRUE,
  market_model = "sim",
  estimation_method = "ols",
  estimation_start = as.Date("2019-01-01"),
  estimation_end = as.Date("2020-02-15")
)

pt <- parametric_tests(list_of_returns = securities_returns,
                 event_start = as.Date("2020-02-16"),
                 event_end = as.Date("2020-03-23")) 

options(digits = 3)
pt <- pt[,c(-2,-3,-4)]
colnames(pt) <- c("Date", "BW80","sig","BW85","sig","t-stat","sig","pt","sig","BH","sig","LMB","sig")
kable(pt,caption = "Table 2: The event study reuslts for the parametric tests")


```


---

# Parametric tests


```{r, echo=FALSE, message=FALSE, warning=FALSE}
securities_returns <- apply_market_model(
  rates = rates,
  regressor = rates_indx,
  same_regressor_for_all = TRUE,
  market_model = "sim",
  estimation_method = "ols",
  estimation_start = as.Date("2019-01-01"),
  estimation_end = as.Date("2020-02-15")
)

pt <- parametric_tests(list_of_returns = securities_returns,
                 event_start = as.Date("2020-02-16"),
                 event_end = as.Date("2020-03-23")) 

options(digits = 3)
pt <- pt[,c(-2,-3,-4)]
colnames(pt) <- c("Date", "BW80","sig","BW85","sig","t-stat","sig","pt","sig","BH","sig","LMB","sig")
kable(pt,caption = "Table 2: The event study reuslts for the parametric tests")


```


---

# Non-parametric tests


```{r, echo=FALSE, message=FALSE, warning=FALSE}
nt <- nonparametric_tests(list_of_returns = securities_returns,
                    event_start = as.Date("2020-02-16"),
                    event_end = as.Date("2020-03-23"))

options(digits = 3)
nt <- nt[,c(-2,-3)]
colnames(nt) <- c("Date", "SIGN","sig","GSIGN","sig","CSIGN","sig","RANK","sig","MRANK","sig","WLCX","sig")
kable(nt,caption = "Table 2: The event study reuslts for the nonparametric tests")
```



---


# First COVID-19 incidence in Italy


```{r vizNOexter, echo=FALSE, message=FALSE, warning=FALSE}
SplitD <- data.frame(name=colnames(rates),
                     when = rep(as.Date("2020-02-21"),12))

Returns <- na.omit(rates)

es1 <- eventstudy(firm.returns = Returns,
                 event.list = SplitD,
                 event.window = 5,
                 type = "None",
                 to.remap = TRUE,
                 remap = "cumsum",
                 inference = TRUE,
                 inference.strategy = "bootstrap")


result1 <- as.data.frame(es1$result) %>% 
  mutate(EventDate = seq(-4,5,1)) %>% 
  rename(ci1 = "2.5%",ci2 = "97.5%")

gg1 <- ggplot(result1,aes(x = EventDate)) +
  geom_line(aes(y=Mean,)) +
  geom_line(aes(y=ci1), linetype="twodash") + 
  geom_line(aes(y=ci2), linetype="twodash") +
  geom_vline(xintercept=0) +
  geom_hline(yintercept = 0) + 
  scale_y_continuous(name="Cum change in response series %") +
  scale_x_continuous(name ="Event Day",labels=c("-4", "-2", "0", "2","4")) +
  theme_bw() +
  ggtitle("")


```

```{r vizNOexter1, echo=FALSE, message=FALSE, warning=FALSE,fig.height= 3}
SplitD <- data.frame(name=colnames(rates),
                     when = rep(as.Date("2020-03-11"),12))

Returns <- na.omit(rates)


es2 <- eventstudy(firm.returns = Returns,
                 event.list = SplitD,
                 event.window = 5,
                 type = "None",
                 to.remap = TRUE,
                 remap = "cumsum",
                 inference = TRUE,
                 inference.strategy = "bootstrap")

result2 <- as.data.frame(es2$result) %>% 
  mutate(EventDate = seq(-4,5,1)) %>% 
  rename(ci1 = "2.5%",ci2 = "97.5%")

gg2 <- ggplot(result2,aes(x = EventDate)) +
  geom_line(aes(y=Mean,)) +
  geom_line(aes(y=ci1), linetype="twodash") + 
  geom_line(aes(y=ci2), linetype="twodash") +
  geom_vline(xintercept=0) +
#  geom_hline(yintercept = 0) + 
  scale_y_continuous(name="Cum change in response series %") +
  scale_x_continuous(name ="Event Day",labels=c("-4", "-2", "0", "2","4")) +
  theme_bw() +
  ggtitle("")




gg1+gg2


```



---

# Local lockdown announcement



```{r vizCROBEXexter, echo=FALSE, message=FALSE, warning=FALSE}

SplitD <- data.frame(name=colnames(rates),
                     when = rep(as.Date("2020-02-21"),12))


es3 <- eventstudy(firm.returns = Returns,
                 event.list = SplitD,
                 event.window = 5,
                 type = "marketModel",
                 to.remap = TRUE,
                 remap = "cumsum",
                 inference = TRUE,
                 inference.strategy = "bootstrap",
                 model.args = list(market.returns=rates_indx$Crobex))

result3 <- as.data.frame(es3$result) %>% 
  mutate(EventDate = seq(-4,5,1)) %>% 
  rename(ci1 = "2.5%",ci2 = "97.5%")

gg3 <- ggplot(result3,aes(x = EventDate)) +
  geom_line(aes(y=Mean,)) +
  geom_line(aes(y=ci1), linetype="twodash") + 
  geom_line(aes(y=ci2), linetype="twodash") +
  geom_vline(xintercept=0) +
  geom_hline(yintercept = 0) + 
  scale_y_continuous(name="Cum change in response series %") +
  scale_x_continuous(name ="Event Day",labels=c("-4", "-2", "0", "2","4")) +
  theme_bw() +
  ggtitle("")

```

```{r vizCROBEXexter2, echo=FALSE, message=FALSE, warning=FALSE,fig.height= 3}

SplitDn <- data.frame(name=colnames(rates),
                     when = rep(as.Date("2020-03-11"),12))


es4 <- eventstudy(firm.returns = Returns,
                 event.list = SplitDn,
                 event.window = 5,
                 type = "marketModel",
                 to.remap = TRUE,
                 remap = "cumsum",
                 inference = TRUE,
                 inference.strategy = "bootstrap",
                 model.args = list(market.returns=rates_indx$Crobex))

result4 <- as.data.frame(es4$result) %>% 
  mutate(EventDate = seq(-4,5,1)) %>% 
  rename(ci1 = "2.5%",ci2 = "97.5%")

gg4 <- ggplot(result4,aes(x = EventDate)) +
  geom_line(aes(y=Mean,)) +
  geom_line(aes(y=ci1), linetype="twodash") + 
  geom_line(aes(y=ci2), linetype="twodash") +
  geom_vline(xintercept=0) +
  geom_hline(yintercept = 0) + 
  scale_y_continuous(name="Cum change in response series %") +
  scale_x_continuous(name ="Event Day",labels=c("-4", "-2", "0", "2","4")) +
  theme_bw() +
  ggtitle("")

gg3 + gg4

```

```{r vizGOLDexter,eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
es3 <- eventstudy(firm.returns = Returns,
                  event.list = SplitD,
                  event.window = 5,
                  type = "marketModel",
                  to.remap = TRUE,
                  remap = "cumsum",
                  inference = TRUE,
                  inference.strategy = "bootstrap",
                  model.args = list(market.returns=rates_indx$Crobex,
                  others=add_indx$gold,
                  market.returns.purge=TRUE))

plot(es3)
```

---

# Summary of findings

- Negative effect of COVID-19 on tourist sector

- Similar market wide effects of COVDID-19 pandemic

- Important events are:
1. The first COVID-19 incidence in Italy
2. Local announcement of lockdown

- How do you explain relative performance of tourist stocks? 

---

class: inverse, center, middle

# Thank you for your attention!

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>























